---
title: "Base usage of DODO"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Base usage of DODO}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, eval = F}
library(DODO)
```

Connect to the neo4j instance.

```{r, eval = F}
connect_to_dodo(local = T)
```

# Basic information on the Neo4j instance.

Show the data model of DODO

```{r, eval = F}
show_dodo_model()
```

List node types and databases that are included. 

```{r, eval = F}
list_database()
list_node_type()
```

```{r, eval = F}
get_ontology("MONDO")
```

# Cross-reference relations

Identify direct cross-reference relations for (an) identifier(s).

```{r, eval = F}
convert_concept(
  ids = "MONDO:0005027",
  relationship = "xref",
  direct = T
)
```

Identify indirect cross-reference relations for (an) identifier(s).

```{r, eval = F}
convert_concept(
  ids = "MONDO:0005027",
  relationship = "xref",
  direct = F
)
```

Identify all indirect parent-child relations for (an) identifier(s).

```{r, eval = F}
convert_concept(
  ids = "MONDO:0005027",
  relationship = "parent", # "child"
  direct = F
)
```

Identify all disease-phenotype or inverse for (an) identifier(s).

```{r, eval = F}
convert_concept(
  ids = "HP:0001250",
  relationship = "disease"
) # "phenotype"
```

Identify alternative identifiers.

```{r, eval = F}
convert_concept(
  ids = "DOID:9489",
  relationship = "alternative"
) # "phenotype"
```

# Plot

Plot the indirect cross-reference relations for (an) identifier(s).

```{r, eval = F}
plot_relations(
  ids = "MONDO:0005027",
  relationship = "xref",
  direct = F
)
```

